BarnebyLives: an R package to create herbarium specimen labels and clean spreadsheets

### 2025-04-23 manuscript/BarnebyLives An R package to create and qc herbarium labels.rmd
scored veg spell check

--- 
title: 'BarnebyLives: an R package to create herbarium specimen labels and clean spreadsheets'
author:  |
    |  $^1$Chicago Botanic Garden, 1000 Lake Cook Road, Glencoe, Illinois 60022, USA  
    |  $^2$Plant Biology and Conservation, Northwestern University, Evanston, Illinois 60208, USA  
    | Reed Clark Benkendorf$^1$$^,$$^2$^[Author for Correspondence: rbenkendorf@chicagobotanic.org], Jeremie B. Fant$^1$$^,$$^2$
abstract: |
  \noindent
  **Premise:** Depositing specimens in herbaria is a time-consuming process. Many institutions have reduced the amount of funding for herbaria and universities have reduced the amount of education dedicated to collections and curatorial tasks. However, the continual generation of herbarium specimens is essential for current and future research on evolution and ecology. To facilitate the continued growth of herbaria, BarnebyLives was developed as a tool to supplement collection notes, perform geographic and taxonomic information processing, enact spell checks and other QC steps, produce labels, and submit digital data to increase the rate of accessioning specimens.  
  **Methods and Results:** BarnebyLives uses geospatial data from the U.S. The Census Bureau provides administrative jurisdictional information and data from other sources, including the United States Geological Survey, to supplement collection notes by providing information on abiotic site conditions. It uses in-house spell checks to verify the spelling of a collection at all taxonomic ranks, the IPNI standard author database to check standard author abbreviations, and the Royal Botanic Garden Kews 'Plants of the World Online' to check for nomenclature innovations. Optionally, the package writes driving directions to the sites using Google Maps. The package outputs data in tabular and spatial formats for review by the user before rendering labels using LaTeX.        
  **Conclusions:** BarnebyLives provides accurate political and physical information, reduces typos, provides users with the most current taxonomic opinions, generates driving directions to sites, and produces aesthetically appealing labels and shipping manifests in a matter of minutes.       
keywords: |
  herbarium, natural history museum, collections, geospatial, R, LaTeX   
output:
  pdf_document: default
  toc: no
  word_document: default
csl: "../citations/american-journal-of-botany.csl"
bibliography: ../citations/citations.bib
link-citations: yes
fig_caption: yes
always_allow_html: yes
header-includes:
- \usepackage{endfloat}
- \usepackage{setspace}\doublespacing
- \usepackage{lineno}
- \linenumbers
- \usepackage[width=\textwidth]{caption}
- \usepackage{wrapfig}
- \usepackage[export]{adjustbox}
- \usepackage{rotating}
---

@@

## Usage
All steps of BarnebyLives, except for label generation are run within the freely available Rstudio. 
Data may be read from any common spreadsheet management system or database connection such as Excel, or free alternatives such as LibreOffice, OpenOffice, or via the cloud on Google Sheets. 
The latter two options are documented here and in package vignettes, detailed descriptions of the required and suggested input columns are located on a Github page (https://github.com/sagesteppe/BarnebyLives 'Input Data Column Names') and 96 real-world examples are on a Google Sheets accessible from the page. 
BarnebyLives is atypical for R packages in that it requires a considerable amount of data to operate (Table 1). 
Virtually all on-disk memory associated with the package are used to store spatial data.
The amount of spatial data varies according to the domain that the user decides to support (Figure 3).
Functions that require on-disk data require a path to data as an argument.  Manually supplying the path argument allows users to judiciously determine a storage location suitable for their needs.

@@

The package was released into beta testing using the primary authors collections from 2023. 
The testing of the package within this manuscript was performed using a subset of their collections from 2018-2022, all of which are un-accessioned. 
Only collections which had identifications to the level of species or lower, and transcribed collection dates and coordinates were used. 
Resulting in a data set of `r nrow(data)` records for testing, from `r n_sites` sites located across Western North America (Figure 2). 
In total this data set had `r n_spp` species (with `r n_sp_authors` distinct sets of authors), with `r n_infraspecies` infraspecies (`r n_infra_sp_auths` authorships) in `r n_families` families were used for testing.

@@

Most of the local run time is attributable to the spatial (`r sum(tt_local[tt_local$Module=='Geospatial', 'elapsed'])`sec), and taxonomic operations (`r sum(tt_local[tt_local$Module=='Taxonomic', 'elapsed'])`sec), while formatting data for labels took `r sum(tt_local[tt_local$Module=='Style', 'elapsed'])`sec. 
The spell check of the scientific name accounted for nearly all of the time (`r sum(tt_local[tt_local$Function=='spell_check', 'elapsed'])`sec) spent performing local taxonomic operations.
The generation of labels consumed around `r minutemen(time_label_gen)` minutes (`r time_label_gen`sec) for the rendering, and an additional `r time_final`sec to combine the `r length(list.files('labels/processed'))` sheets to a single Portable Document Format (PDF). 
The total computer run time for processing these `r nrow(data)` specimens was `r minutemen(sum(time_trials$elapsed))` minutes.

@@

## Family Level Results
# these species were found by POW, and according to POW, the species are in the f
# following families. 
family_incongruence <- data_p %>% 
  sf::st_drop_geometry() %>% 
  filter(POW_Family != 'NOT FOUND') %>% 
  filter(POW_Family != Family) %>% 
  select(Collection_number, Scientific_name, Family, POW_Family) %>% 
  mutate(Situation = 
    case_when(
      Family %in% c('Hydrophyllaceae', 'Namaceae') ~ 'Preferred', 
      Family %in% c('Athyriaceae') ~ 'Outdated',
      Family %in% c('Saxifragaceae', 'Plantaginaceae') ~ 'Submitter_Incorrect', 
      Family %in% c('Bataceae', 'Ericaceae') ~ 'Internal_Error', 
      .default = 'Typo'
    ))

@@

# the mismatch between POW genus and not
genus_mismatch <- data_p %>% 
  sf::st_drop_geometry() %>% 
  filter(POW_Genus != 'NOT FOUND') %>% 
  filter(POW_Genus != Genus) %>% 
  select(Collection_number, Scientific_name, Genus, POW_Genus) %>% 
  mutate(Situation = 
    case_when(
      Genus %in% c('Glandularia', 'Oenothera', 'Chamerion') ~ 'Preferred',
      # check Pleuraphis, i am confused on this still....
      Genus %in% c('Cryptantha', 'Ivesia', 'Lotus', 'Polygala', 'Minuartia', 'Pleuraphis',
                    'Pseudostellaria', 'Achnatherum', 'Scirpus', 'Peritoma', 'Arenaria', 
                    'Spartina', 'Polygonum', 'Atriplex', 'Cymopterus', 'Trisetum', 'Bahia', 
                   'Oryzopsis', 'Potentilla', 'Pascopyrum'
                    ) ~ 'Outdated',
      Genus %in% c('Vaccinium', 'Eleocharis') ~ 'Internal_Error',
      .default = 'Typo'
    ),
    Situation = if_else(Scientific_name == 'Oenothera caespitosa Nutt.', 'Internal_Error', Situation))

@@

epithet_diff <- data_p %>% 
  sf::st_drop_geometry() %>% 
  filter(Epithet != POW_Epithet, !Scientific_name %in% genus_mismatch$Scientific_name) %>% 
  select(Collection_number, Scientific_name, Epithet, POW_Genus, POW_Epithet, POW_Infraspecies) %>% 
  distinct(Scientific_name, .keep_all = T) %>%  # do we record repeat collections ?
  mutate(Situation = 
           case_when(
             Collection_number %in% c(
               '1217', '1251', '1283', '1311', '1337', '1391', '1415', '1421', '1435', '1450', '1576', 
               '1594', '1624', '1627', '1745', '2043', '2086', '2108', '2109', '2120', '2127', '2130', '2135', 
               '2153', '2199', '2227', '2242', '2256', '2289', '2297', '2315', '2343', '2370', '2379', 
               '2393', '2415', '2480', '2511', '2561', '2573', '2648', '2660', 
               '2687', '2690', '2710', '2715', '2727', '2750', '2767', '2818')
             ~ 'Spelling',
             Collection_number %in% c(
               '1556', '1588', '1619', '2445', '2777', '1337', '2004', '2412', '2487', '2748', '2652',
               '2805', '2194', '2040', '2039', '2403', '2328', '2076') 
             ~ 'Outdated', 
             Collection_number %in% c('2644') ~ 'Incorrect',
             Collection_number %in% c(
               '1207', '2326', '2038', '1424', '2618', '2622', '2665', '2711', '2713') 
             ~ 'Internal_Error',
             Collection_number %in% c('2110', '2540') ~ 'Preferred'
           )
  )

rm(Family_in, Family_out)
```

@@

Of the `r nrow(data)` records analysed (`r n_spp` distinct species) BarnebyLives flagged `r nrow(epithet_diff)` records, and detected `r sum(epithet_diff$Situation=='Spelling')`  instances of misspelled epithets (`r sum(epithet_diff$Situation=='Spelling')` unique species). 
Of the `r nrow(data)` species analysed (`r n_spp` distinct species) BarnebyLives flagged `r nrow(epithet_diff)` records, and detected `r sum(epithet_diff$Situation=='Spelling')`  instances of misspelled epithets (`r sum(epithet_diff$Situation=='Spelling')` unique species). 
In `r sum(epithet_diff$Situation=='Outdated')` of these instances the user supplied an outdated name (`r nrow(unique(epithet_diff[epithet_diff$Situation=='Outdated', 'Scientific_name']))` unique species). 
It also flagged `r sum(epithet_diff$Situation=='Preferred', na.rm = T)` records where the author follows an alternative taxonomy (`r nrow(unique(epithet_diff[epithet_diff$Situation=='Preferred', 'Scientific_name']))` unique species), and flagged `r sum(epithet_diff$Situation=='Internal_Error', na.rm = T)` records in error. 
The final record was an egregious error where the order of the specific epithet and the genus name.

`r long_flag` records were appropriately flagged for issues with auto fill increment of the longitude value, and `r lat_flag` records were also auto-flagged for increases in latitude values (% of records).

@@

# ORCID
Reed Clark Benkendorf https://orcid.org/0000-0003-3110-6687  
Jeremie Fant https://orcid.org/0000-0001-9276-1111

>>>>

--- 
title: 'BarnebyLives: an R package to create herbarium specimen labels and clean spreadsheets'
author:  |
    |  $^1$Chicago Botanic Garden, 1000 Lake Cook Road, Glencoe, Illinois 60022, USA  
    |  $^2$Plant Biology and Conservation, Northwestern University, Evanston, Illinois 60208, USA  
    | Reed Clark Benkendorf$^1$$^,$$^2$^[Author for Correspondence: rbenkendorf@chicagobotanic.org], Jeremie B. Fant$^1$$^,$$^2$
abstract: |
  \noindent
  **Premise:** Depositing specimens in herbaria is a time-consuming process. Many institutions have reduced the amount of funding for herbaria and universities have reduced the amount of education dedicated to collections and curatorial tasks. However, the continual generation of herbarium specimens is essential for current and future research on evolution and ecology. To facilitate the continued growth of herbaria, BarnebyLives was developed as a tool to supplement collection notes, perform geographic and taxonomic information processing, enact spell checks and other QC steps, produce labels, and submit digital data to increase the rate of accessioning specimens.  
  **Methods and Results:** BarnebyLives uses geospatial data from the U.S. The Census Bureau provides administrative jurisdictional information and data from other sources, including the United States Geological Survey, to supplement collection notes by providing information on abiotic site conditions. It uses in-house spell checks to verify the spelling of a collection at all taxonomic ranks, the IPNI standard author database to check standard author abbreviations, and the Royal Botanic Garden Kews 'Plants of the World Online' to check for nomenclature innovations. Optionally, the package writes driving directions to the sites using Google Maps. The package outputs data in tabular and spatial formats for review by the user before rendering labels using LaTeX.        
  **Conclusions:** BarnebyLives provides accurate political and physical information, reduces typos, provides users with the most current taxonomic opinions, generates driving directions to sites, and produces aesthetically appealing labels and shipping manifests in a matter of minutes.       
keywords: |
  herbarium, natural history museum, collections, geospatial, R, LaTeX   
output:
  pdf_document: default
  toc: no
  word_document: default
csl: "../citations/american-journal-of-botany.csl"
bibliography: ../citations/citations.bib
link-citations: yes
fig_caption: yes
always_allow_html: yes
header-includes:
  - \usepackage{endfloat}
  - \usepackage{setspace}\doublespacing
  - \usepackage{lineno}
  - \linenumbers
  - \usepackage{rotating}
---

@@

## Usage
All steps of BarnebyLives, except for label generation are run within the freely available Rstudio. 
Data may be read from any common spreadsheet management system or database connection such as Excel, or free alternatives such as LibreOffice, OpenOffice, or via the cloud on Google Sheets. 
The latter two options are documented here and in package vignettes, detailed descriptions of the required and suggested input columns are located on a Github page (https://github.com/sagesteppe/BarnebyLives 'Input Data Column Names') and 96 real-world examples are on a Google Sheets accessible from the page. 
BarnebyLives is atypical for R packages in that it requires a considerable amount of data to operate (Table 1). 
Virtually all on-disk memory associated with the package are used to store spatial data.
The amount of spatial data varies according to the domain that the user decides to support (Figure 3).
Functions that require on-disk data require a path to data as an argument.  Manually supplying the path argument allows users to determine an appropriate storage location suitable for their needs.

@@

The package was released into beta testing using the primary authors collections from 2023. 
The testing of the package within this manuscript was performed using a subset of their collections from 2018-2022. 
Only collections which had identifications to the level of species or lower, and transcribed collection dates and coordinates were used. 
Resulting in a data set of `r nrow(data)` records for testing, from `r n_sites` sites located across Western North America (Figure 2). 
In total this data set had `r n_spp` species (with `r n_sp_authors` distinct sets of authors), with `r n_infraspecies` infraspecies (`r n_infra_sp_auths` authorships) in `r n_families` families which were used for testing.

@@

Most of the local run time is attributable to the spatial (`r sum(tt_local[tt_local$Module=='Geospatial', 'elapsed'])`sec), and taxonomic operations (`r sum(tt_local[tt_local$Module=='Taxonomic', 'elapsed'])`sec), while formatting data for labels took `r sum(tt_local[tt_local$Module=='Style', 'elapsed'])`sec. 
The spell check of the scientific name accounted for nearly all of the time (`r sum(tt_local[tt_local$Function=='spell_check', 'elapsed'])`sec) spent performing local taxonomic operations.
The generation of labels consumed around `r minutemen(time_label_gen)` minutes (`r time_label_gen`sec) for the rendering, and an additional `r time_final`sec to combine the `r length(list.files('labels/processed'))` **sheets** to a single Portable Document Format (PDF). 
The total computer run time for processing these `r nrow(data)` specimens was `r minutemen(sum(time_trials$elapsed))` minutes.

@@

## Family Level Results
# these species were found by POW, and according to POW, the species are in the f
# following families. 
family_incongruence <- data_p %>% 
  sf::st_drop_geometry() %>% 
  filter(POW_Family != 'NOT FOUND') %>% 
  filter(POW_Family != Family) %>% 
  select(Collection_number, Scientific_name, Family, POW_Family) %>% 
  mutate(Situation = 
    case_when(
      Family %in% c('Hydrophyllaceae', 'Namaceae') ~ 'Preferred', 
      Family %in% c('Athyriaceae') ~ 'Outdated',
      Family %in% c('Saxifragaceae', 'Plantaginaceae') ~ 'Submitter_Incorrect', 
      Family %in% c('Ranunculaceae', 'Ericaceae') ~ 'Internal_Error', 
      .default = 'Typo'
    ))

@@

# the mismatch between POW genus and not
genus_mismatch <- data_p %>% 
  sf::st_drop_geometry() %>% 
  filter(POW_Genus != 'NOT FOUND') %>% 
  filter(POW_Genus != Genus) %>% 
  select(Collection_number, Scientific_name, Genus, POW_Genus) %>% 
  mutate(Situation = 
    case_when(
      Genus %in% c('Glandularia', 'Oenothera', 'Chamerion') ~ 'Preferred',
      # check Pleuraphis, i am confused on this still....
      Genus %in% c('Cryptantha', 'Ivesia', 'Lotus', 'Polygala', 'Minuartia', 'Pleuraphis',
                    'Pseudostellaria', 'Achnatherum', 'Scirpus', 'Peritoma', 'Arenaria', 
                    'Spartina', 'Polygonum', 'Atriplex', 'Cymopterus', 'Trisetum', 'Bahia', 
                   'Oryzopsis', 'Potentilla', 'Pascopyrum'
                    ) ~ 'Outdated',
      Genus %in% c('Allium', 'Vaccinium', 'Eleocharis') ~ 'Internal_Error',
      .default = 'Typo'
    ),
    Situation = if_else(Scientific_name == 'Oenothera caespitosa Nutt.', 'Internal_Error', Situation))

@@

epithet_diff <- data_p %>% 
  sf::st_drop_geometry() %>% 
  filter(Epithet != POW_Epithet, !Scientific_name %in% genus_mismatch$Scientific_name) %>% 
  select(Collection_number, Scientific_name, Epithet, POW_Genus, POW_Epithet, POW_Infraspecies) %>% 
  distinct(Scientific_name, .keep_all = T) %>%  # do we record repeat collections ?
  mutate(Situation = 
           case_when(
             Collection_number %in% c('1207') ~ 'User Error', 
             Collection_number %in% c(
               '1217', '1251', '1283', '1311', '1391', '1415', '1435', '1450', '1556','1576', 
               '1594', '1624', '1627', '1745', '2043', '2108', '2109', '2120', '2127', '2130', '2135', 
               '2153', '2199', '2227', '2242', '2256', '2289', '2297', '2315', '2370', '2379', 
               '2393', '2480', '2511', '2561', '2573', '2648', '2660', 
               '2687', '2690', '2710', '2715', '2727', '2750', '2767', '2818')
             ~ 'Spelling',
             Collection_number %in% c(
               '1337', '1421', '1588', '1619','2415', '2445', '2777', '1337', '2004', '2086', '2343', '2412', '2487', '2748', '2652',
               '2805', '2194', '2040', '2039', '2403',  '2076') 
             ~ 'Outdated', 
             Collection_number %in% c('2644') ~ 'Incorrect',
             Collection_number %in% c(
               '1207', '2326', '2038', '1424', '2618', '2622', '2665', '2711', '2713') 
             ~ 'Internal_Error',
             Collection_number %in% c('2110', '2252', '2328', '2540') ~ 'Preferred'
           )
  )

rm(Family_in, Family_out)

## check spell checks on both vegetation and associates columns

veg_diff <- data_p %>% 
  sf::st_drop_geometry() %>% 
  select(Collection_number, Vegetation, SpellCk.Vegetation) %>% 
  filter(Vegetation != SpellCk.Vegetation) %>% 
  distinct(Vegetation, SpellCk.Vegetation, .keep_all = TRUE) %>%

  # grouped sites - these sites are essentially replicates of each other, I have just
  # remove a focal taxon from them. I feel they are pseudoreplicates so they will be discarded

  filter(!Collection_number %in%
   c('2058', '2107', '2110', '2112', '2117', '2220', '2221', '2407', '2428', '2429', 
     '2430', '2432', '2540', '2541', '2560', '1407') 
  ) %>% 
  mutate(Situation = 
           case_when(
             Collection_number %in% c(  
               '1207', '1286', '1596', '1598', '2004', '2033', '2034', '2133', 
               '2175', '2402', '2407', '2427', '2760' 
               ) ~ 'User Error',
             Collection_number %in% c(
               '1214', '1218', '1220', '1227', '1228', '1250', '1257', '1258', '1270', 
               '1273', '1287', '1292', '1293', '1302', '1341', '1343', '1383', 
               '1384', '1390', '1401', '1455', '1557', '1558', '1565', '1596', 
               '2000', '2010', '2012', '2014', '2032', '2049', '2054', '2067', '2105', 
               '2116', '2163', '2219', '2224', '2230', '2395', '2406', '2454', '2528', 
               '2539', '2562', '2566', '2571', '2572', '2622', '2625', '2656', 
               '2664', '2668', '2745', '2759', '2787'
             )
             ~ 'Spelling',
             Collection_number %in% c(
                '1278', '1339', '1405'
               ) ~ 'Incorrect', # that is the program returns the wrong taxon
             Collection_number %in% c(
               '2527' 
             ) 
             ~ 'Internal_Error',
           )
  )
```

@@

Of the `r nrow(data)` records analysed (`r n_spp` distinct species) BarnebyLives flagged `r nrow(epithet_diff)` records, and detected `r sum(epithet_diff$Situation=='Spelling')`  instances of misspelled epithets (`r sum(epithet_diff$Situation=='Spelling')` unique species). 
Of the `r nrow(data)` species analysed (`r n_spp` distinct species) BarnebyLives flagged `r nrow(epithet_diff)` records, and detected `r sum(epithet_diff$Situation=='Spelling')`  instances of misspelled epithets (`r sum(epithet_diff$Situation=='Spelling')` unique species). 
In `r sum(epithet_diff$Situation=='Outdated')` of these instances the user supplied an outdated name (`r nrow(unique(epithet_diff[epithet_diff$Situation=='Outdated', 'Scientific_name']))` unique species). 
It also flagged `r sum(epithet_diff$Situation=='Preferred', na.rm = T)` records where the author follows an alternative taxonomy (`r nrow(unique(epithet_diff[epithet_diff$Situation=='Preferred', 'Scientific_name']))` unique species), and flagged `r sum(epithet_diff$Situation=='Internal_Error', na.rm = T)` record in error. 
The final record was an egregious error where the order of the specific epithet and the genus name.

`r long_flag` records were appropriately flagged for issues with auto fill increment of the longitude value, and `r lat_flag` records were also auto-flagged for increases in latitude values (% of records). 
Vegetation data were entered for `r nrow(data) - sum(is.na(data$Vegetation))` records, and after removal of duplicate site information BL flagged  `r nrow(veg_diff)` records. 
BL was able to correct the spelling in `r nrow(veg_diff[veg_diff$Situation == 'Spelling',])` instances, and in  `r nrow(veg_diff[veg_diff$Situation == 'User Error',])` or instances the user entered data in a format beyond the capabilities of BL to fix. 
BL returned an incorrect spell check match for  `r nrow(veg_diff[veg_diff$Situation == 'Incorrect',])` records, two of which had the same misspelled species, in `r nrow(veg_diff[veg_diff$Situation == 'Internal_Error',])` instance BL had an internal errors where it returned an incorrect match due to abbreviations.

@@

\singlespacing
# ORCID
Reed Clark Benkendorf https://orcid.org/0000-0003-3110-6687  
Jeremie Fant https://orcid.org/0000-0001-9276-1111
